---
title: "Progress Memo 1"
subtitle: |
  | Final Project 
  | Data Science 1 with R (STAT 301-1)
author: "Oscar Baliga"
pagetitle: "PM1 Oscar Baliga"
date: today

format:
  html:
    toc: true
    toc-depth: 4
    toc-location: left
    embed-resources: true
    link-external-newwindow: true
    code_folding: hide

execute:
  warning: false

from: markdown+emoji
reference-location: margin
citation-location: margin
---

::: {.callout-tip icon="false"}
## Github Repo Link

[Oscar's Link](https://github.com/oscbal/final_project.git)
:::

```{r}
#| echo: FALSE
#Load Packages
library(skimr)
library(knitr)
library(rmarkdown)
library(tidyverse)
library(naniar)
library(maps)
library(plotly)

#Load data
bomber_data <- read.csv("data/operations.csv")
```

## Introduction to the Data

### Data Source

This dataset consists of digitized paper mission reports from World War II, developed by Lt Col Jenns Robertson of the US Air Force. The records include data from U.S. and Royal Air Force missions, as well as contributions from Australian, New Zealand, and South African air forces. The dataset captures a live-action sequence of air warfare from 1939 to 1945[^1]

[^1]: [Link to Dataset. Accessed on 11/05/24.](https://www.kaggle.com/datasets/usaf/world-war-ii?resource=download)

### Why This Data

As a double major in Data Science and History, I am fascinated by the intersection of these two fields. Analyzing historical data from significant events like World War II allows me to engage with history from a new perspective. The digitized mission reports provide opportunities to uncover patterns that traditional methods might overlook. I can investigate how factors such as altitude and aircraft type impacted mission outcomes. This approach not only enhances my understanding of military history but also develops my data analysis skills. Ultimately, I aim to provide new insights into World War II that can inform both academic and public discourse.

### Data Quality & Complexity Check

Before conducting any analysis, a quality check on the dataset will be necessary. Key aspects include:

```         
- **Completeness**: Identifying if any critical information, such as mission dates or locations, is missing. 

- **Consistency**: Checking for uniformity in data entries, particularly in conflict names and geographic locations. 

- **Validity**: Ensuring that the recorded dates and mission outcomes conform to expected formats and categories. 

- **Complexity**: The dataset includes various data elements that may require careful handling to ensure accurate analysis, especially when merging with other datasets or conducting time series analyses.
```

### Potential Data Issues

Several potential issues may affect the usability of this dataset:

\-**Missing Values:** Some records lack details. All character values have a 100% complete rate, but the numeric data possess issues with completeness:

```         
Source ID: Complete rate of 0.023 (97.7% missing).

Target Latitude: Complete rate of 0.049 (95.1% missing).

Target Longitude: Complete rate of 0.049 (95.1% missing).

Total Weight (Tons): Complete rate of 0.066 (93.4% missing).

High Explosives Weight (Tons): Complete rate of 0.18 (82% missing).

Attacking Aircraft: Complete rate of 0.298 (70.2% missing).

Altitude (Hundreds of Feet): Complete rate of 0.322 (67.8% missing).
```

-   **Data Entry Errors**: There might be inconsistencies in how missions and conflicts are labeled or recorded, leading to difficulties in categorization and analysis. Especially with datasets from the 1940s, it must be taken into consideration that decades of clerical errors may have altered the data.

-   **Outliers**: There are many outliers for most numerical variables in the dataset:

    Takeoff Longitude: 429 outliers

```{r}
#| label: dist_1
#| fig-cap: ""
#| echo: FALSE
#| code-fold: true

bomber_data |>
  ggplot(aes(x = Takeoff.Longitude)) +
  geom_boxplot() +
  labs(
    x = "Takeoff Longitude",
    title = "Distribution of Takeoff Longitudes for Allied Bombers in WW2"
  ) +
  theme_light()
```

```         
Target Latitude: 145 outliers


Target Longitude: 127 outliers
```

```{r}
#| label: dist_3
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = Target.Longitude)) +
  geom_boxplot() +
  labs(
    x = "Target Latitude",
    title = "Distribution of Target Latitude for Allied Bombers in WW2"
  ) +
  theme_light()
```

```         
Altitude Hundreds of Feet: 29
```

```{r}
#| label: dist_4
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = Target.Latitude)) +
  geom_boxplot() +
  labs(
    x = "Target Latitude",
    title = "Distribution of Target Latitude for Allied Bombers in WW2"
  ) +
  theme_light()
```

```         
Airborne Aircraft: 1823
```

```{r}
#| label: dist_5
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = Airborne.Aircraft)) +
  geom_boxplot() +
  labs(
    x = "Airborne Aircraft",
    title = "Distribution of Airborne Aircraft for Allied Bombers in WW2"
  ) +
  theme_light()
```

```         
Attacking Aircraft: 9601
```

```{r}
#| label: dist_6
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = Takeoff.Longitude)) +
  geom_boxplot() +
  labs(
    x = "Takeoff Longitude",
    title = "Distribution of Takeoff Longitudes for Allied Bombers in WW2"
  ) +
  theme_light()
```

```         
Bombing Aircraft: 1417
```

```{r}
#| label: dist_7
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = Bombing.Aircraft)) +
  geom_boxplot() +
  labs(
    x = "Bombing Aircraft",
    title = "Distribution of Bombing.Aircraft for Allied Bombers in WW2"
  ) +
  theme_light()
```

```         
Aircraft Returned: 2
```

```{r}
#| label: dist_8
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = Aircraft.Returned)) +
  geom_boxplot() +
  labs(
    x = "Aircraft Returned",
    title = "Distribution of Aircraft Returned for Allied Bombers in WW2"
  ) +
  theme_light()
```

```         
Aircraft Failed: 32
```

```{r}
#| label: dist_9
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = Aircraft.Failed)) +
  geom_boxplot() +
  labs(
    x = "Aircraft Failed",
    title = "Distribution of Aircraft Failed for Allied Bombers in WW2"
  ) +
  theme_light()
```

```         
Aircraft Damaged: 2
```

```{r}
#| label: dist_10
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = Aircraft.Damaged)) +
  geom_boxplot() +
  labs(
    x = "Aircraft Damaged",
    title = "Distribution of Aircraft Damaged for Allied Bombers in WW2"
  ) +
  theme_light()
```

```         
Aircraft Lost: 18
```

```{r}
#| label: dist_11
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = Aircraft.Lost)) +
  geom_boxplot() +
  labs(
    x = "Aircraft Lost",
    title = "Distribution of Aircraft Lost for Allied Bombers in WW2"
  ) +
  theme_light()
```

```         
High Explosives: 3537
```

```{r}
#| label: dist_12
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = High.Explosives)) +
  geom_boxplot() +
  labs(
    x = "High Explosives",
    title = "Distribution of High Explosives for Allied Bombers in WW2"
  ) +
  theme_light()
```

```         
High Explosives Weight Tons: 7568
```

```{r}
#| label: dist_13
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = High.Explosives.Weight..Tons.)) +
  geom_boxplot() +
  labs(
    x = "High Explosives in Weight of Tons",
    title = "Distribution of High Explosives in Weight of Tons for Allied Bombers in WW2"
  ) +
  theme_light()
```

```         
Incendiary Devices: 515
```

```{r}
#| label: dist_14
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = Incendiary.Devices)) +
  geom_boxplot() +
  labs(
    x = "Incendiary Devices",
    title = "Distribution of Incendiary Devices for Allied Bombers in WW2"
  ) +
  theme_light()
```

```         
Incendiary Devices Weight Pounds: 69
```

```{r}
#| label: dist_15
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = Incendiary.Devices.Weight..Pounds.)) +
  geom_boxplot() +
  labs(
    x = "Incendiary Devices in Weight of Pounds",
    title = "Distribution of Incendiary Devices in Weight of Pounds for Allied Bombers in WW2"
  ) +
  theme_light()
```

```         
Incendiary Devices Weight Tons: 2081
```

```{r}
#| label: dist_16
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = Incendiary.Devices.Weight..Tons.)) +
  geom_boxplot() +
  labs(
    x = "Incendiary Devices in Weight of Tons",
    title = "Distribution of Incendiary Devices in Weight of Tons for Allied Bombers in WW2"
  ) +
  theme_light()
```

```         
Fragmentation Devices: 457
```

```{r}
#| label: dist_17
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = Fragmentation.Devices)) +
  geom_boxplot() +
  labs(
    x = "Fragmentation Devices",
    title = "Distribution of Fragmentation Devices for Allied Bombers in WW2"
  ) +
  theme_light()
```

```         
Fragmentation Devices Weight Pounds: 14
```

```{r}
#| label: dist_18
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = Fragmentation.Devices.Weight..Pounds.)) +
  geom_boxplot() +
  labs(
    x = "Fragmentation Devices in Weight of Pounds",
    title = "Distribution of Fragmentation Devices in Weight of Pounds for Allied Bombers in WW2"
  ) +
  theme_light()
```

```         
Fragmentation Devices Weight Tons: 2284
```

```{r}
#| label: dist_19
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = Fragmentation.Devices.Weight..Tons.)) +
  geom_boxplot() +
  labs(
    x = "Fragmentation Devices in Weight of Tons",
    title = "Distribution of Fragmentation Devices in Weight of Tons for Allied Bombers in WW2"
  ) +
  theme_light()
```

```         
Total Weight Pounds: 136
```

```{r}
#| label: dist_20
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = Total.Weight..Pounds.)) +
  geom_boxplot() +
  labs(
    x = "Total Weight in Pounds",
    title = "Distribution of Total Weight in Pounds for Allied Bombers in WW2"
  ) +
  theme_light()
```

```         
Total Weight Tons: 9363
```

```{r}
#| label: dist_21
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = Total.Weight..Tons.)) +
  geom_boxplot() +
  labs(
    x = "Total Weight in Tons",
    title = "Distribution of Total Weight in Tons for Allied Bombers in WW2"
  ) +
  theme_light()
```

-   **Bias in Data Collection**: The dataset may reflect biases based on the documentation practices of the time or the focus on certain operations over others. Furthermore, many record of this age may have been lost/damaged during combat or during the period between the end of WW2 and digitization of the dataset.

### Misc

The dataset offers a unique opportunity to understand the operational history of air forces during WWII. It serves not only as a historical record but also as a potential source for analyzing broader themes such as technological advancements in warfare and international military collaboration.

## Reasearch

### Guiding Question:

How did allied bombing operations differ between different time periods, air forces, theaters, and tactical strategies.

### Overarall Distribution of Bombing activities during the War

This map showcases the cumulative bombing damage wrought upon the Axis Powers by observing the locations bombed by various Allied air forces. The dots vary in size by the weight of bombs dropped and vary in color by which country the attack belonged to.

```{r}
#| label: dist_22
#| fig-cap: ""
#| echo: FALSE

bomber_data_filtered_by_long_lat <- bomber_data %>%
  mutate(Country = ifelse(is.na(Country) | Country == "", "NOT STATED", Country)) |>
  filter(
    between(Target.Latitude, -90, 90) &
    between(Target.Longitude, -180, 180)
  )


  ggplot() +
    geom_polygon(data = map_data("world"), aes(x = long, y = lat, group = group), fill = "gray80") +
    geom_point(
      data = bomber_data_filtered_by_long_lat, 
      aes(
        x = Target.Longitude, 
        y = Target.Latitude, 
        color = Country, 
        size = Total.Weight..Tons.), 
      alpha = 0.01) +
    coord_map() +
    labs(
      title = "All Allied Locations of Bombardment during WW2",
      x = "Longitude",
      y = "Latitude",
      color = "Country",
      size = "Tonnage of Munitions per Bombardment"
    ) +
    theme_light()
  
```

Somewhat overloaded with information, this map truly portrays the magnitude of the Second World War. With dots scattered across 4 continents, the amount of resources needed to sustain just one side of the war must have been truly enormous. Looking at the Easter Theater, it's apparent that the United States played the dominate role militarily in the East. Due to MacArthur's island hopping campaign, pink dots are scattered across the Eastern Pacific as the fortified Japanese Army was gradually pried from their conquests. In the Western Theater, we can see a far greater level of international cooperation. This can partially be explained by the Royal Air Force being active in the area since 1939 (the United States only joined the war in '41) and the bombardment tactic of the Royal Air Force bombing German targets during the night and the Americans bombing Germany during the day, creating relative equilibrium. Furthermore, the greater presence of bombardments without nationalities (under "NOT STATED") in the West may support the claim of greater international cooperation in the West because the air force command structure between countries may have been so integrated they didn't deem it important to specify which group or combinations of groups took action.

### Number of Bombardements by Country

```{r}
#| label: dist_23
#| fig-cap: ""
#| echo: FALSE
bomber_data_with_not_stated <- bomber_data |>
    mutate(Country = ifelse(is.na(Country) | Country == "", "NOT STATED", Country))


bomber_data_with_not_stated |>
  ggplot(aes(y = Country, fill = Country)) +
  geom_bar()
```
To answer the earlier questions raised of the ratio of bombardments by country, this graph shows that the US is by far in the lead. Interestingly, "Not Stated" is second by a wide margin to the UK. This likely means that many of the "Not Stated" bombardments were carried through by the Royal Air Force and may not have been documented due to the UK being to only nation fighting for a period of time or the international aspects of the British Armed services from the Commonwealth or exiled governments in the UK and using RAF resources.

### Bombardments by Country Over Time

```{r}
#| label: dist_24
#| fig-cap: ""
#| echo: FALSE
bomber_data_with_not_stated <- bomber_data |>
    mutate(Country = ifelse(is.na(Country) | Country == "", "NOT STATED", Country),
           Year = year(mdy(Mission.Date)))


bomber_data_with_not_stated |>
  ggplot(aes(y = Country, fill = Country)) +
  geom_bar() +
  facet_wrap(~Year) +
  scale_x_continuous(
    breaks = seq(0, 50000, 10000), 
    labels = seq(0, 5, 1) 
  ) +
  labs(
    title = "Number of Bombardments by Country Over Time",
    x ="Number"
  )
```
In these graphs we can observe the number of bombardments by country. Based upon the graph, the earlier claim that the "Not Stated" bombers were predominantly British is likely incorrect. Early in the war when the UK was the only Allied Power, there were no "Not Stated" values for countries, but, when America enters the war, the number of "USA" bombardments grows at the same rate as the "Not Stated" bombardments. This correlation in rate of growth is a strong sign that many of the "Not Stated" bombardments are predominantly American, not British.


### Differences in Bombardment for Japan And Germany

##### Top 10 Targeted Industries by East vs. West
###### Western Theater
```{r} 
#| label: dist_25
#| fig-cap: ""
#| echo: FALSE

European_Indus_Targets <- bomber_data |>
  filter(Theater.of.Operations %in% c("ETO", "MTO")) |>
  count(Target.Industry) |>
  arrange(desc(n)) |>
  slice(1:10) |>
  pull(Target.Industry)

kable(European_Indus_Targets)
```

###### Eastern Theater
```{r} 
#| label: dist_26
#| fig-cap: ""
#| echo: FALSE

Asian_Indus_Targets <- bomber_data |>
  filter(Theater.of.Operations %in% c("PTO", "CBI")) |>
  count(Target.Industry) |>
  arrange(desc(n)) |>
  slice(1:10) |>
  pull(Target.Industry)

kable(Asian_Indus_Targets)
```

Here we can see a stark difference in the record keeping between the Eastern and Western Theaters of combat. In the West, Industry targeting data was kept very well, even if it was a civilian target, while in the East target industry was not noted. This difference could be for various reasons, but it was likely not due to Air forces not strategically targeting different industries in the Eastern Theater because that would be a very large strategical oversight. 

**More will be done in this section**

### Closer Look  at Germany


#### Germany Overall
```{r}
#| label: dist_27
#| fig-cap: ""
#| echo: FALSE

germany_bombed <- bomber_data |>
  filter(Target.Country == "GERMANY") |>
  filter(
    between(Target.Latitude, 45, 57) &
    between(Target.Longitude, 0, 20)
  )



ggplot() +
  geom_polygon(data = map_data("world", region = "Germany"), aes(x = long, y = lat, group = group), fill = "gray80") +
  geom_point(
    data = germany_bombed, 
    aes(
      x = Target.Longitude, 
      y = Target.Latitude,
      color = Country,
      size = Total.Weight..Tons.), 
    alpha = 0.01) +
  coord_map() +
  labs(
    title = "Allied Locations of Bombardment during WW2 in Germany",
    x = "Longitude",
    y = "Latitude",
    color = "Country",
    size = "Tonnage of Munitions per Bombardment"
  ) +
  theme_light()
```

In this graph we can observe the distribution of bombing attacks across Germany. From the faint dots to the East of modern Germany, we can see areas that once were within the legal boundaries of Germany during this time but were removed post-war. We can also observe the focus on the industrial sector in the Ruhr Valley where many coal plants were located. 

#### Looking at Berlin

```{r}
#| label: dist_28
#| fig-cap: ""
#| echo: FALSE
berlin_bombed <- germany_bombed |>
  filter(
    between(Target.Latitude, 52, 53) &
      between(Target.Longitude, 12.75, 14.25)
  )

top_7_industries <- berlin_bombed |>
  count(Target.Industry) |>
  arrange(desc(n)) |>
  head(7) |>
  pull(Target.Industry)

berlin_bombed_top7 <- berlin_bombed |>
  filter(Target.Industry %in% top_7_industries)

ggplot() +
  geom_polygon(data = map_data("world", region = "Germany"), aes(x = long, y = lat, group = group), fill = "gray80") +
  geom_point(
    data = berlin_bombed_top7,
    aes(
      x = Target.Longitude,
      y = Target.Latitude,
      color = Target.Industry,
      size = Total.Weight..Tons.
    ),
    alpha = 0.08
  ) +
  coord_map(xlim = c(12.75, 14.25), ylim = c(52, 53)) + 
  labs(
    title = "Allied Locations of Bombardment during WW2 in Berlin",
    x = "Longitude",
    y = "Latitude",
    color = "Target Industry",
    size = "Tonnage of Munitions per Bombardment"
  ) +
  theme_light()
```

From this graph we can get a closer look to the strategies used for bombardment in urban areas. In this about 100 km by 100 km box much of Germany's population, government, and industry resided. Around the city center we can see many green dot, signifying many urban targets, but can also see railroad targets, due to city centers of 20th century European cities having extensive transportation infrastructure. On the edges of the city, military targets encircle the city, a sign the defensive lines that protected the heart of Nazi Germany. Seeing the amount of bombings done in the area, show cases how dangerous life was for civilians in wartime Germany and the pressures put upon the German government to sustain infrastructure through these difficult times.

### Closer look at Japan
```{r}
japan_bombed <- bomber_data |>
  filter(Target.Country == "JAPAN") |>
  filter(
    between(Target.Latitude, 20, 45) &
    between(Target.Longitude, 122, 153)
  )



ggplot() +
  geom_polygon(data = map_data("world", region = "Japan"), aes(x = long, y = lat, group = group), fill = "gray80") +
  geom_point(
    data = japan_bombed, 
    aes(
      x = Target.Longitude, 
      y = Target.Latitude,
      color = Incendiary.Devices,
      size = Total.Weight..Tons.), 
    alpha = 0.01) +
  coord_map() +
  labs(
    title = "Allied Locations of Bombardment during WW2 in Japan",
    x = "Longitude",
    y = "Latitude",
    color = "Country",
    size = "Tonnage of Munitions per Bombardment"
  ) +
  theme_light()
```
