---
title: "Progress Memo 1"
subtitle: |
  | Final Project 
  | Data Science 1 with R (STAT 301-1)
author: "Oscar Baliga"
pagetitle: "PM1 Oscar Baliga"

format:
  html:
    toc: true
    toc-depth: 4
    toc-location: left
    embed-resources: true
    link-external-newwindow: true
    code_folding: hide

execute:
  warning: false

from: markdown+emoji
reference-location: margin
citation-location: margin
---


::: {.callout-tip icon=false}
## Date Created: 11/2/2024
:::

::: {.callout-tip icon=false}

## Github Repo Link

[Oscar's Link](https://github.com/stat301-1-2024-fall/final-project-1-oscbal.git)

:::

```{r}
#| echo: FALSE
#Load Packages
library(skimr)
library(knitr)
library(rmarkdown)
library(tidyverse)
library(naniar)

#Load data
bomber_data <- read.csv("data/operations.csv")
```

## Data Source

This dataset consists of digitized paper mission reports from World War II, developed by Lt Col Jenns Robertson of the US Air Force. The records include data from U.S. and Royal Air Force missions, as well as contributions from Australian, New Zealand, and South African air forces. The dataset captures a live-action sequence of air warfare from 1939 to 1945^[This dataset can be found on Kaggle at the following link: https://www.kaggle.com/datasets/usaf/world-war-ii?resource=download].

##Why This Data

As a double major in Data Science and History, I am fascinated by the intersection of these two fields. Analyzing historical data from significant events like World War II allows me to engage with history from a new perspective. The digitized mission reports provide opportunities to uncover patterns that traditional methods might overlook. I can investigate how factors such as altitude and aircraft type impacted mission outcomes. This approach not only enhances my understanding of military history but also develops my data analysis skills. Ultimately, I aim to provide new insights into World War II that can inform both academic and public discourse.

## Data Quality & Complexity Check

Before conducting any analysis, a quality check on the dataset will be necessary. Key aspects include:
- **Completeness**: Identifying if any critical information, such as mission dates or locations, is missing.
- **Consistency**: Checking for uniformity in data entries, particularly in conflict names and geographic locations.
- **Validity**: Ensuring that the recorded dates and mission outcomes conform to expected formats and categories.
- **Complexity**: The dataset includes various data elements that may require careful handling to ensure accurate analysis, especially when merging with other datasets or conducting time series analyses.

## Potential Data Issues

Several potential issues may affect the usability of this dataset:

-**Missing Values: **Some records lack details. All character values have a 100% complete rate, but the numeric data possess issues with completeness:

    Source ID: Complete rate of 0.023 (97.7% missing).

    Target Latitude: Complete rate of 0.049 (95.1% missing).

    Target Longitude: Complete rate of 0.049 (95.1% missing).

    Total Weight (Tons): Complete rate of 0.066 (93.4% missing).

    High Explosives Weight (Tons): Complete rate of 0.18 (82% missing).

    Attacking Aircraft: Complete rate of 0.298 (70.2% missing).

    Altitude (Hundreds of Feet): Complete rate of 0.322 (67.8% missing).

- **Data Entry Errors**: There might be inconsistencies in how missions and conflicts are labeled or recorded, leading to difficulties in categorization and analysis. Especially with datasets from the 1940s, it must be taken into consideration that decades of clerical errors may have altered the data.

- **Outliers**: There are many outliers for most numerical variables in the dataset:


    Takeoff Longitude: 429 outliers
    
```{r}
#| label: dist_1
#| fig-cap: ""
#| echo: FALSE
#| code-fold: true

bomber_data |>
  ggplot(aes(x = Takeoff.Longitude)) +
  geom_boxplot() +
  labs(
    x = "Takeoff Longitude",
    title = "Distribution of Takeoff Longitudes for Allied Bombers in WW2"
  ) +
  theme_light()
```
    

    Target Latitude: 145 outliers
    

    Target Longitude: 127 outliers
    
```{r}
#| label: dist_3
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = Target.Longitude)) +
  geom_boxplot() +
  labs(
    x = "Target Latitude",
    title = "Distribution of Target Latitude for Allied Bombers in WW2"
  ) +
  theme_light()
```

    Altitude Hundreds of Feet: 29
    
```{r}
#| label: dist_4
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = Target.Latitude)) +
  geom_boxplot() +
  labs(
    x = "Target Latitude",
    title = "Distribution of Target Latitude for Allied Bombers in WW2"
  ) +
  theme_light()
```

    Airborne Aircraft: 1823
    
```{r}
#| label: dist_5
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = Airborne.Aircraft)) +
  geom_boxplot() +
  labs(
    x = "Airborne Aircraft",
    title = "Distribution of Airborne Aircraft for Allied Bombers in WW2"
  ) +
  theme_light()
```

    Attacking Aircraft: 9601
    
```{r}
#| label: dist_6
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = Takeoff.Longitude)) +
  geom_boxplot() +
  labs(
    x = "Takeoff Longitude",
    title = "Distribution of Takeoff Longitudes for Allied Bombers in WW2"
  ) +
  theme_light()
```

    Bombing Aircraft: 1417
    
```{r}
#| label: dist_7
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = Bombing.Aircraft)) +
  geom_boxplot() +
  labs(
    x = "Bombing Aircraft",
    title = "Distribution of Bombing.Aircraft for Allied Bombers in WW2"
  ) +
  theme_light()
```

    Aircraft Returned: 2
    
```{r}
#| label: dist_8
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = Aircraft.Returned)) +
  geom_boxplot() +
  labs(
    x = "Aircraft Returned",
    title = "Distribution of Aircraft Returned for Allied Bombers in WW2"
  ) +
  theme_light()
```

    Aircraft Failed: 32
  
```{r}
#| label: dist_9
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = Aircraft.Failed)) +
  geom_boxplot() +
  labs(
    x = "Aircraft Failed",
    title = "Distribution of Aircraft Failed for Allied Bombers in WW2"
  ) +
  theme_light()
```

    Aircraft Damaged: 2
    
```{r}
#| label: dist_10
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = Aircraft.Damaged)) +
  geom_boxplot() +
  labs(
    x = "Aircraft Damaged",
    title = "Distribution of Aircraft Damaged for Allied Bombers in WW2"
  ) +
  theme_light()
```

    Aircraft Lost: 18

```{r}
#| label: dist_11
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = Aircraft.Lost)) +
  geom_boxplot() +
  labs(
    x = "Aircraft Lost",
    title = "Distribution of Aircraft Lost for Allied Bombers in WW2"
  ) +
  theme_light()
```

    High Explosives: 3537
    
```{r}
#| label: dist_12
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = High.Explosives)) +
  geom_boxplot() +
  labs(
    x = "High Explosives",
    title = "Distribution of High Explosives for Allied Bombers in WW2"
  ) +
  theme_light()
```

    High Explosives Weight Tons: 7568
    
```{r}
#| label: dist_13
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = High.Explosives.Weight..Tons.)) +
  geom_boxplot() +
  labs(
    x = "High Explosives in Weight of Tons",
    title = "Distribution of High Explosives in Weight of Tons for Allied Bombers in WW2"
  ) +
  theme_light()
```

    Incendiary Devices: 515
    
```{r}
#| label: dist_14
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = Incendiary.Devices)) +
  geom_boxplot() +
  labs(
    x = "Incendiary Devices",
    title = "Distribution of Incendiary Devices for Allied Bombers in WW2"
  ) +
  theme_light()
```

    Incendiary Devices Weight Pounds: 69
    
```{r}
#| label: dist_15
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = Incendiary.Devices.Weight..Pounds.)) +
  geom_boxplot() +
  labs(
    x = "Incendiary Devices in Weight of Pounds",
    title = "Distribution of Incendiary Devices in Weight of Pounds for Allied Bombers in WW2"
  ) +
  theme_light()
```

    Incendiary Devices Weight Tons: 2081
    
```{r}
#| label: dist_16
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = Incendiary.Devices.Weight..Tons.)) +
  geom_boxplot() +
  labs(
    x = "Incendiary Devices in Weight of Tons",
    title = "Distribution of Incendiary Devices in Weight of Tons for Allied Bombers in WW2"
  ) +
  theme_light()
```

    Fragmentation Devices: 457
    
```{r}
#| label: dist_17
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = Fragmentation.Devices)) +
  geom_boxplot() +
  labs(
    x = "Fragmentation Devices",
    title = "Distribution of Fragmentation Devices for Allied Bombers in WW2"
  ) +
  theme_light()
```

    Fragmentation Devices Weight Pounds: 14
    
```{r}
#| label: dist_18
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = Fragmentation.Devices.Weight..Pounds.)) +
  geom_boxplot() +
  labs(
    x = "Fragmentation Devices in Weight of Pounds",
    title = "Distribution of Fragmentation Devices in Weight of Pounds for Allied Bombers in WW2"
  ) +
  theme_light()
```

    Fragmentation Devices Weight Tons: 2284
    
```{r}
#| label: dist_19
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = Fragmentation.Devices.Weight..Tons.)) +
  geom_boxplot() +
  labs(
    x = "Fragmentation Devices in Weight of Tons",
    title = "Distribution of Fragmentation Devices in Weight of Tons for Allied Bombers in WW2"
  ) +
  theme_light()
```

    Total Weight Pounds: 136
    
```{r}
#| label: dist_20
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = Total.Weight..Pounds.)) +
  geom_boxplot() +
  labs(
    x = "Total Weight in Pounds",
    title = "Distribution of Total Weight in Pounds for Allied Bombers in WW2"
  ) +
  theme_light()
```

    Total Weight Tons: 9363
    
```{r}
#| label: dist_21
#| fig-cap: ""
#| echo: FALSE

bomber_data |>
  ggplot(aes(x = Total.Weight..Tons.)) +
  geom_boxplot() +
  labs(
    x = "Total Weight in Tons",
    title = "Distribution of Total Weight in Tons for Allied Bombers in WW2"
  ) +
  theme_light()
```




- **Bias in Data Collection**: The dataset may reflect biases based on the documentation practices of the time or the focus on certain operations over others. Furthermore, many record of this age may have been lost/damaged during combat or during the period between the end of WW2 and digitization of the dataset.

## Misc

The dataset offers a unique opportunity to understand the operational history of air forces during WWII. It serves not only as a historical record but also as a potential source for analyzing broader themes such as technological advancements in warfare and international military collaboration.

