---
title: "Final Project: WW2 Bombing Study"
subtitle: |
  | Final Project 
  | Data Science 1 with R (STAT 301-1)
author: "Oscar Baliga"
pagetitle: "PM1 Oscar Baliga"
date: today

format:
  html:
    toc: true
    toc-depth: 4
    toc-location: left
    embed-resources: true
    link-external-newwindow: true
    code_folding: hide

execute:
  warning: false

from: markdown+emoji
reference-location: margin
citation-location: margin
---

::: {.callout-tip icon="false"}
## Github Repo Link

[Oscar's Link](https://github.com/oscbal/final_project.git)
:::

```{r}
#| echo: FALSE
#Load Packages
library(skimr)
library(knitr)
library(rmarkdown)
library(tidyverse)
library(naniar)
library(maps)
library(plotly)

#Load data
bomber_data <- read.csv("data/operations.csv")
```

## Introduction to the Data

### Data Source

This dataset consists of digitized paper mission reports from World War II, developed by Lt Col Jenns Robertson of the US Air Force. The records include data from U.S. and Royal Air Force missions, as well as contributions from Australian, New Zealand, and South African air forces. The dataset captures a live-action sequence of air warfare from 1939 to 1945[^1]

[^1]: [Link to Dataset. Accessed on 11/05/24.](https://www.kaggle.com/datasets/usaf/world-war-ii?resource=download)

### Why This Data

As a double major in Data Science and History, I am fascinated by the intersection of these two fields. Analyzing historical data from significant events like World War II allows me to engage with history from a new perspective. The digitized mission reports provide opportunities to uncover patterns that traditional methods might overlook. I can investigate how factors such as altitude and aircraft type impacted mission outcomes. This approach not only enhances my understanding of military history but also develops my data analysis skills. Ultimately, I aim to provide new insights into World War II that can inform both academic and public discourse.

### Data Quality & Complexity Check

Before conducting any analysis, a quality check on the dataset will be necessary. Key aspects include:

```         
- Completeness: Identifying if any critical information, such as mission dates or locations, is missing. 

- Consistency: Checking for uniformity in data entries, particularly in conflict names and geographic locations. 

- Validity: Ensuring that the recorded dates and mission outcomes conform to expected formats and categories. 

- Complexity: The dataset includes various data elements that may require careful handling to ensure accurate analysis, especially when merging with other datasets or conducting time series analyses.
```

### Potential Data Issues

Several potential issues may affect the usability of this dataset:

\-**Missing Values:** Some records lack details. All character values have a 100% complete rate, but the numeric data possess issues with completeness:

```         
Source ID: Complete rate of 0.023 (97.7% missing).

Target Latitude: Complete rate of 0.049 (95.1% missing).

Target Longitude: Complete rate of 0.049 (95.1% missing).

Total Weight (Tons): Complete rate of 0.066 (93.4% missing).

High Explosives Weight (Tons): Complete rate of 0.18 (82% missing).

Attacking Aircraft: Complete rate of 0.298 (70.2% missing).

Altitude (Hundreds of Feet): Complete rate of 0.322 (67.8% missing).
```

-   **Data Entry Errors**: There might be inconsistencies in how missions and conflicts are labeled or recorded, leading to difficulties in categorization and analysis. Especially with datasets from the 1940s, it must be taken into consideration that decades of clerical errors may have altered the data.

-   **Outliers**: There are many outliers for most numerical variables in the dataset:
skim_wihtoutcharts
```{r}
#| label: dist_0
#| fig-cap: ""
#| echo: FALSE
#| code-fold: true

date_overview <- bomber_data |>
  mutate(across(where(is.character), ~ na_if(., ""))) |>
  skimr::skim_without_charts()

kable(date_overview)
```


    Takeoff Longitude: 429 outliers

```         
Target Latitude: 145 outliers


Target Longitude: 127 outliers
```

```         
Altitude Hundreds of Feet: 29
```

```         
Airborne Aircraft: 1823
```

```         
Attacking Aircraft: 9601
```

```         
Bombing Aircraft: 1417
```


```         
Aircraft Returned: 2
```

```         
Aircraft Failed: 32
```

```         
Aircraft Damaged: 2
```

```         
Aircraft Lost: 18
```

```         
High Explosives: 3537
```

```         
High Explosives Weight Tons: 7568
```

```         
Incendiary Devices: 515
```

```         
Incendiary Devices Weight Pounds: 69
```

```         
Incendiary Devices Weight Tons: 2081
```

```         
Fragmentation Devices: 457
```

```         
Fragmentation Devices Weight Pounds: 14
```

```         
Fragmentation Devices Weight Tons: 2284
```

```         
Total Weight Pounds: 136
```

```         
Total Weight Tons: 9363
```

-   **Bias in Data Collection**: The dataset may reflect biases based on the documentation practices of the time or the focus on certain operations over others. Furthermore, many record of this age may have been lost/damaged during combat or during the period between the end of WW2 and digitization of the dataset.

### Misc

The dataset offers a unique opportunity to understand the operational history of air forces during WWII. It serves not only as a historical record but also as a potential source for analyzing broader themes such as technological advancements in warfare and international military collaboration.

## Reasearch

### Guiding Question:

How did allied bombing operations differ between different time periods, air forces, theaters, and tactical strategies.

### Overarall Distribution of Bombing activities during the War

This map showcases the cumulative bombing damage wrought upon the Axis Powers by observing the locations bombed by various Allied air forces. The dots vary in size by the weight of bombs dropped and vary in color by which country the attack belonged to.

```{r}
#| label: dist_22
#| fig-cap: ""
#| echo: FALSE

bomber_data_filtered_by_long_lat <- bomber_data |>
  mutate(Country = ifelse(is.na(Country) | Country == "", "NOT STATED", Country)) |>
  filter(
    between(Target.Latitude, -90, 90) &
    between(Target.Longitude, -180, 180)
  )


  ggplot() +
    geom_polygon(data = map_data("world"), aes(x = long, y = lat, group = group), fill = "gray80") +
    geom_point(
      data = bomber_data_filtered_by_long_lat, 
      aes(
        x = Target.Longitude, 
        y = Target.Latitude, 
        color = Country, 
        size = Total.Weight..Tons.), 
      alpha = 0.01) +
    coord_map() +
    labs(
      title = " All Allied Locations of Bombardment \n during WW2",
      x = "Longitude",
      y = "Latitude",
      color = "Country",
      size = "Tonnage of Munitions \nper Bombardment"
    ) +
    theme_light() +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.key.size = unit(.3, "cm")
  ) +
    guides(
    color = guide_legend(override.aes = list(alpha = 1)),
    size = guide_legend(override.aes = list(alpha = 1))
  )
  
```

Somewhat overloaded with information, this map truly portrays the magnitude of the Second World War. With dots scattered across 4 continents, the amount of resources needed to sustain just one side of the war must have been truly enormous. Looking at the Easter Theater, it's apparent that the United States played the dominate role militarily in the East. Due to MacArthur's island hopping campaign, pink dots are scattered across the Eastern Pacific as the fortified Japanese Army was gradually pried from their conquests. In the Western Theater, we can see a far greater level of international cooperation. This can partially be explained by the Royal Air Force being active in the area since 1939 (the United States only joined the war in '41) and the bombardment tactic of the Royal Air Force bombing German targets during the night and the Americans bombing Germany during the day, creating relative equilibrium. Furthermore, the greater presence of bombardments without nationalities (under "NOT STATED") in the West may support the claim of greater international cooperation in the West because the air force command structure between countries may have been so integrated they didn't deem it important to specify which group or combinations of groups took action.


### All Locations of departure for Allied Aircaft

```{r}
#| label: dist_22.5
#| fig-cap: ""
#| echo: FALSE

bomber_data_filtered_by_long_lat <- bomber_data |>
  mutate(Country = ifelse(is.na(Country) | Country == "", "NOT STATED", Country)) |>
   mutate(
    Takeoff.Latitude = as.numeric(Takeoff.Latitude),
    Takeoff.Longitude = as.numeric(Takeoff.Longitude)
  ) |>
  filter(
  !is.na(Takeoff.Latitude) & !is.na(Takeoff.Longitude) &
  between(Takeoff.Latitude, -90, 90) &
  between(Takeoff.Longitude, -180, 180)
  )


  ggplot() +
    geom_polygon(data = map_data("world"), aes(x = long, y = lat, group = group), fill = "gray80") +
    geom_point(
      data = bomber_data_filtered_by_long_lat, 
      aes(
        x = Takeoff.Longitude, 
        y = Takeoff.Latitude, 
        color = Country, 
        size = Total.Weight..Tons.), 
      alpha = 0.01) +
    coord_map() +
    labs(
      title = " All Allied Locations of takeoff \n during WW2",
      x = "Longitude",
      y = "Latitude",
      color = "Country",
      size = "Tonnage of Munitions \nper Bombardment"
    ) +
    theme_light() +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.key.size = unit(.3, "cm")
  ) +
    guides(
    color = guide_legend(override.aes = list(alpha = 1)),
    size = guide_legend(override.aes = list(alpha = 1))
  )
  
```

This graph shows where Allied bombers took off during WW2, broken down by country and the tonnage of munitions. The missingness rate is extremely high, which makes it hard to take the data at face value. That said, the points that are included probably line up with areas where the battle lines were really fluid, like North Africa or Southeast Asia, so they were more likely to be recorded. Places like England, which were more stable but integral for bombing Germany and Western Europe, might not show up here because of how the data was collected. The bases near the frontlines or major targets probably got more attention in the records since they were so important for operations. Bottom line: this shows some interesting patterns, but the missing data makes it tricky to draw any solid conclusions.

### Number of Bombardements by Country

```{r}
#| label: dist_23
#| fig-cap: ""
#| echo: FALSE
bomber_data_with_not_stated <- bomber_data |>
    mutate(Country = ifelse(is.na(Country) | Country == "", "NOT STATED", Country))


bomber_data_with_not_stated |>
  ggplot(aes(y = Country, fill = Country)) +
  geom_bar()
```
To answer the earlier questions raised of the ratio of bombardments by country, this graph shows that the US is by far in the lead. Interestingly, "Not Stated" is second by a wide margin to the UK. This likely means that many of the "Not Stated" bombardments were carried through by the Royal Air Force and may not have been documented due to the UK being to only nation fighting for a period of time or the international aspects of the British Armed services from the Commonwealth or exiled governments in the UK and using RAF resources.

### Bombardments by Country Over Time

```{r}
#| label: dist_24
#| fig-cap: ""
#| echo: FALSE
bomber_data_with_not_stated <- bomber_data |>
    mutate(Country = ifelse(is.na(Country) | Country == "", "NOT STATED", Country),
           Year = year(mdy(Mission.Date)))


bomber_data_with_not_stated |>
  ggplot(aes(y = Country, fill = Country)) +
  geom_bar() +
  facet_wrap(~Year) +
  scale_x_continuous(
    breaks = seq(0, 50000, 10000), 
    labels = seq(0, 5, 1) 
  ) +
  labs(
    title = "Number of Bombardments by Country Over Time",
    x ="Count in 10,000s"
  )
```
In these graphs we can observe the number of bombardments by country. Based upon the graph, the earlier claim that the "Not Stated" bombers were predominantly British is likely incorrect. Early in the war when the UK was the only Allied Power, there were no "Not Stated" values for countries, but, when America enters the war, the number of "USA" bombardments grows at the same rate as the "Not Stated" bombardments. This correlation in rate of growth is a strong sign that many of the "Not Stated" bombardments are predominantly American, not British.

### Aircraft Used By Year
```{r}
#| label: dist_24.5
#| fig-cap: ""
#| echo: FALSE

top_aircraft_types <- bomber_data %>%
  mutate(
    Mission.Date = mdy(Mission.Date),  
    year = year(Mission.Date)       
  ) %>%
  group_by(year, Aircraft.Series) %>% 
  summarize(n = n(), .groups = "drop") %>% 
  group_by(year) %>% 
  slice_max(n, n = 7) %>% 
  ungroup()
  

ggplot(top_aircraft_types, aes(x = reorder(Aircraft.Series, n), y = n, fill = as.factor(year))) +
  geom_col() +
  coord_flip() +  # Flip coordinates for better readability
  facet_wrap(~ year, scales = "free_y") +  # Create facets for each year
  labs(
    title = "Top 10 Aircraft Types Used in Missions by Year",
    x = "Aircraft Type",
    y = "Number of Missions",
    fill = "Year"
  ) +
  theme_minimal()
  
```
This chart shows how aircraft technology and usage evolved throughout World War II. Early in the war (1939-1941), British planes like the Wellington ("WELL") and Hampden ("HAMP") were heavily used, reflecting Britain’s early involvement and reliance on bombers like the Blenheim ("BLEN") for operations in Europe. By 1942, American aircraft such as the B-17 Flying Fortress and B-24 Liberator started to dominate, showing the U.S.'s growing role in the war and their focus on building heavy bombers with greater range and payload capacity. From 1943 onward, planes like the B-17, B-24, and the B-26 Marauder became essential, with advanced models like the P-51 Mustang appearing by 1945 for escort and bombing missions. This shift highlights how Allied aircraft became more advanced and versatile as the war progressed.

### Differences in Bombardment for Japan And Germany

##### Top 10 Targeted Industries by East vs. West
###### Western Theater
```{r} 
#| label: dist_25
#| fig-cap: ""
#| echo: FALSE

European_Indus_Targets <- bomber_data |>
  filter(Theater.of.Operations %in% c("ETO", "MTO")) |>
  count(Target.Industry) |>
  arrange(desc(n)) |>
  slice(1:10) |>
  pull(Target.Industry)

kable(European_Indus_Targets)
```

###### Eastern Theater
```{r} 
#| label: dist_26
#| fig-cap: ""
#| echo: FALSE

Asian_Indus_Targets <- bomber_data |>
  filter(Theater.of.Operations %in% c("PTO", "CBI")) |>
  count(Target.Industry) |>
  arrange(desc(n)) |>
  slice(1:10) |>
  pull(Target.Industry)

kable(Asian_Indus_Targets)
```

Here we can see a stark difference in the record keeping between the Eastern and Western Theaters of combat. In the West, Industry targeting data was kept very well, even if it was a civilian target, while in the East target industry was not noted. This difference could be for various reasons, but it was likely not due to Air forces not strategically targeting different industries in the Eastern Theater because that would be a very large strategical oversight. 

**More will be done in this section**

### Closer Look  at Germany


#### Germany Overall
```{r}
#| label: dist_27
#| fig-cap: ""
#| echo: FALSE

germany_bombed <- bomber_data |>
  filter(Target.Country == "GERMANY") |>
  mutate(Country = ifelse(is.na(Country) | Country == "", "NOT STATED", Country)) |>
  filter(
    between(Target.Latitude, 45, 57) &
    between(Target.Longitude, 0, 20)
  )



ggplot() +
  geom_polygon(data = map_data("world", region = "Germany"), aes(x = long, y = lat, group = group), fill = "gray80") +
  geom_point(
    data = germany_bombed, 
    aes(
      x = Target.Longitude, 
      y = Target.Latitude,
      color = Country,
      size = Total.Weight..Tons.), 
    alpha = 0.01) +
  coord_map() +
  labs(
    title = "Allied Locations of Bombardment during WW2 in Germany",
    x = "Longitude",
    y = "Latitude",
    color = "Country",
    size = "Tonnage of Munitions \nper Bombardment"
  ) +
  theme_light()  +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.key.size = unit(.3, "cm")
  ) +
    guides(
    color = guide_legend(override.aes = list(alpha = 1)),
    size = guide_legend(override.aes = list(alpha = 1))
  )
```

In this graph we can observe the distribution of bombing attacks across Germany. From the faint dots to the East of modern Germany, we can see areas that once were within the legal boundaries of Germany during this time but were removed post-war. We can also observe the focus on the industrial sector in the Ruhr Valley where many coal plants were located. 

#### Looking at Berlin

```{r}
#| label: dist_28
#| fig-cap: ""
#| echo: FALSE
berlin_bombed <- germany_bombed |>
  filter(
    between(Target.Latitude, 52, 53) &
      between(Target.Longitude, 12.75, 14.25)
  )

top_3_industries <- berlin_bombed |>
  count(Target.Industry) |>
  arrange(desc(n)) |>
  head(3) |>
  pull(Target.Industry)

berlin_bombed_top7 <- berlin_bombed |>
  filter(Target.Industry %in% top_3_industries)

ggplot() +
  geom_polygon(data = map_data("world", region = "Germany"), aes(x = long, y = lat, group = group), fill = "gray80") +
  geom_point(
    data = berlin_bombed_top7,
    aes(
      x = Target.Longitude,
      y = Target.Latitude,
      color = Target.Industry,
      size = Total.Weight..Tons.
    ),
    alpha = 0.08
  ) +
  coord_map(xlim = c(12.75, 14.25), ylim = c(52, 53)) + 
  labs(
    title = "Allied Locations of Bombardment \nduring WW2 in Berlin",
    x = "Longitude",
    y = "Latitude",
    color = "Target Industry",
    size = "Tonnage of Munitions per Bombardment"
  ) +
  theme_light()  +
    guides(
    color = guide_legend(override.aes = list(alpha = 1)),
    size = guide_legend(override.aes = list(alpha = 1))
  )
```

From this graph we can get a closer look to the strategies used for bombardment in urban areas. In this about 100 km by 100 km box much of Germany's population, government, and industry resided. Around the city center we can see many green dot, signifying many urban targets, but can also see railroad targets, due to city centers of 20th century European cities having extensive transportation infrastructure. On the edges of the city, military targets encircle the city, a sign the defensive lines that protected the heart of Nazi Germany. Seeing the amount of bombings done in the area, show cases how dangerous life was for civilians in wartime Germany and the pressures put upon the German government to sustain infrastructure through these difficult times.


### Closer look at Japan

#### Japan Overall
```{r}
#| label: dist_29
#| fig-cap: ""
#| echo: FALSE
japan_bombed <- bomber_data |>
  filter(Target.Country == "JAPAN") |>
  filter(
    between(Target.Latitude, 20, 45) &
    between(Target.Longitude, 122, 153)
  )

ggplot() +
  geom_polygon(data = map_data("world", region = "Japan"), aes(x = long, y = lat, group = group), fill = "gray80") +
  geom_point(
    data = japan_bombed, 
    aes(
      x = Target.Longitude, 
      y = Target.Latitude,
      color = Total.Weight..Tons.), 
    alpha = 0.01) +
  coord_map() +
  labs(
    title = "Allied Locations of Bombardment during WW2 in Japan",
    x = "Longitude",
    y = "Latitude",
    color = "Weight in tons",
  ) +
  scale_color_gradient(
    low = "blue", 
    high = "darkblue", 
    na.value = "gray50"  # Handle missing values, if any
  ) +
  theme_light()
```
From this map we can observe the key locations bombed in Japan. As much of the country is mountainous, industry and civilian targets lie closer to the coast. Furthermore, the two largest cities, Tokyo and Osaka, are clearly visible. Furthermore, Japan's southern island, Kyushu, was heavily targeted with incendiary bombs, as seen by the grouping of dots. Japan's Eastern coast appears to be far less scathed which may be due to the lower population in these areas and the difficulties of traversing the mountains in the center of the islands.

#### Bombs falling on Japan over time
```{r}
#| label: dist_30
#| fig-cap: ""
#| echo: FALSE
japan_bombed_summary <- japan_bombed |>
  mutate(
    Mission.Date = mdy(Mission.Date),           # Convert to Date format
    Month_Year = floor_date(Mission.Date, "month") # Round down to first day of the month
  ) |>
  group_by(Month_Year) |>
  summarise(Total_Tons_Dropped = sum(Total.Weight..Tons., na.rm = TRUE)) # Sum tons dropped

# Plot the aggregated data
ggplot(japan_bombed_summary, aes(x = Month_Year, y = Total_Tons_Dropped)) +
  geom_col(fill = "blue") +  # Use geom_col() for summed data
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
  labs(
    title = "Bombing in Japan by Month and Year",
    x = "Month and Year",
    y = "Total Tons of Bombs Dropped"
  ) +
  theme_light()


```
This graph shows the tonnage of bombs dropped on Japan during World War II, with a sharp increase in 1944 and 1945 reflecting the U.S.'s intensifying bombing campaigns. The early years, including the 1942 Doolittle Raid, are not prominent on the graph due to the limited scale of these operations; the raid's significance was psychological rather than destructive. The turning point came with the capture of key islands like Saipan and Tinian in 1944, enabling B-29 bombers to reach Japan. By 1945, sustained air raids, including the devastating firebombing of Tokyo, marked the peak of the campaign. This escalation was critical in forcing Japan's surrender and demonstrates the importance of geographic and logistical advances in the Pacific theater.
